---
title: "BIOL266_Lab4_Assignment"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
remove(list=ls())
```

In this lab we will be studying the output of *OrthoFinder*, a popular program
that uses local alignment tools like BLAST and distance-based clustering 
approaches to identify sets of homologous genes across species. The input to
OrthoFinder is a set of FASTA files, one for each species, containing 
unaligned protein-coding sequences generated from RNA-Seq or genome assembly
annotations. You can find the documentation of OrthoFinder here: 

https://github.com/davidemms/orthofinder

The primary output of OrthoFinder is a set of *orthogroups*, sets of gene 
families across species that may be related by speciation (orthologs), 
gene duplication and loss (paralogs), or horizontal transfer (xenologs). Since
it is a summary of gene presence/absence variation across multiple genomes, these
datasets are often referred to as *pangenomes*. We will be exploring an orthogroup 
dataset generated for 56 angiosperm species by the Plants and Python NSF REU 
teaching module at Michigan State University. Data source here: 

https://github.com/PlantsAndPython/PlantsAndPython/tree/master/P_9_COMPARATIVE_GENOMICS

We will be working with the data file `Practice9_Orthogroups.GeneCount.csv`, 
which is provided in this week's assignment folder. Load the dataset into
R below: 

```{r}
```

Our goal is to explore patterns in this dataset, with the aim of identifying
a curated list of orthogroups that may be useful for downstream phylogenetic
reconstruction. First, consider the gene count data contained in an individual
orthogroup (row). What does variation in gene count across species tell you
about how these genes are related? For example, if species A has 3 gene copies
and species B has one gene copy for an orthogroup, how would you describe the 
evolutionary history of this orthogroup? Which sets of genes are orthologs and
which are paralogs? 

```
```

What proportion of orthogroups in this dataset contain enough information
to potentially be useful for phylogenetic inference? 

```{r}
```
Plot the number of orthogroups as a function of the number of species (k) that
contain at least one gene copy for that orthogroup. Make two plots: 1) one with
the number of orthogroups containing genes in *exactly* k species, and one with
the number of orthogroups containing genes in *at least* k species. What 
relationships do you see? How many orthogroups contain at least one gene in all 
56 species? What might explain these patterns? (Hint: the relationship in
the exactly k plot is common in pangenome datasets).

```{r}
```
```
```

Recreate the at least k plot, but requiring exactly one gene copy in each species,
rather than at least one (i.e. strict single-copy orthologs). How did the 
results change? Why are we interested in this set of orthogroups for phylogenetic
inference? 

```{r}
```
```
```

We can also explore species-level patterns with this data. Create plots of
1) the total number of genes per species (tallied across all orthogroups); 2)
the mean number of genes per orthogroup per species (excluding zeroes). What
patterns do you notice? What do you think might be contributing to these 
patterns? 

```{r}
```

```
```
In phylogenetics, there is generally a tradeoff between taxonomic sampling (
the number of species represented in each gene alignment) and locus sampling (
the number of genetic loci sampled per species). In general, you want enough
taxonomic representation to be able to infer relationships among all your 
studies species, and enough genetic loci to have statistical confidence in
your inferred relationships among those species. Given these constraints,
what criteria would you apply to the angiosperm dataset to generate a final
list of orthologs for phylogenetic reconstruction? Explain your reasoning
below: 

```
```

