---
title: "BIOL266_Lab5_Assignment"
output: html_document
date: "2026-09-04"
---

```{r}
remove(list=ls())
```

In this lab, we will focus on understanding distance methods for phylogenetic
reconstruction. The folder for this lab assignment contains five multiple 
sequence alignment datasets in fasta format. Open the first file, 
`dataset01_balanced.fasta`, in the space below. There are several R packages 
that contain functions for parsing fasta files, although the data structure 
they are stored in in R can differ.

```{r}
```
From this multiple sequence alignment, calculate two genetic distance matrices,
one using Hamming distance and one using Jukes-Cantor distances.You may use 
external functions for this step. In the Hamming distance matrix, divide by
the number of sites in the alignment (10,000) so it is normalized to per-site
distances.

```{r}
```
Compare the distances values in these two matrices. How are they different, and
why might you be seeing these differences? Which matrix should you use for 
phylogenetic reconstruction?

```
```
In the space below, define two functions that take your constructed genetic
distance matrix as input. One function should implement the UPGMA algorithm
for phylogenetic reconstruction, and the other should implement neighbor-joining.
Both functions should return a `Phylo` class object representing the constructed
phylogeny. For this task, you may use external helper functions for things like
individual calculaations or constructing the `Phylo` object, but each step of
the algorithms themselves should be implemented by you and clearly documented.

```{r}
```
```
```

Using both methods, reconstruct phylogenies from the remaining four provided
multiple sequence alignments, and print the results. 

```{r}
```
What do the differences between the UPGMA and NJ trees tell you about evolutionary
rates in the underlying genetic datasets? 

```
```
