---
title: "BIOL266_Lab2_Assignment"
output:
  pdf_document: default
  html_document: default
date: "2026-05-20"
---

Welcome to RStudio! This document is written in RMarkdown. RStudio comes with
a tool called **knitr** which generates html or pdf documents from your 
RMarkdown file. This can be done by clicking the *Knit* button above the 
code window in RStudio. For your assignments, please submit both the completed
RMarkdown file and an html or pdf file generated from that file. 

R Code in RMarkdown files is embedded in *chunks* with a particular format. 
Here is an example showing some basic math operations you can do in R: 

```{r}
1+3
3-1
3*10^2
10 / pi
sin(4)
log(100) #natural log by default
```

You can run chunks one at a time and test their outputs by pressing the green 
"play" button to the  right of the chunk in RStudio. Chunks will also be run 
automatically when you knit your Rmd file. 

Like with bash, R is a full scripting language with tools like variable
assignment and loops. Some examples: 

```{r}
a <- 10
b <- a + 10

for (i in 1:5) { #Loop with five iterations
  a <- a + 10 #Updating the value assigned to variable a
  print(a) #This prints the value of a
  b <- a + 10
  print(b) #This prints the value of b
}
```

Note that variable assignment in R can be done with `<-` rather than `=` 
(though both work). This actually helps with code clarity because it
helps distinguish variable assignment from mathematical operations. 

When re-running code, it's often useful to clear stored variables from R's 
memory so that their values are what you expect them to be. You can
do this with 

```{r}
rm(list=ls())
```

`ls` prints all currently variables, while `rm` removes them. I usually put this
at the top of my R scripts. 

R is a popular language for data science and statistics, and it has many objects
designed for storing and manipulating data that you will encounter frequently
while completing assignments in this class. *Vectors* are collections of data
of a similar type, typically either numeric (e.g. 17.3), integer (e.g. 4), or
character (e.g. "Bob"). 

To check the type of your data, you can use a *function*; these are objects
that take some input, apply a series of operations to it, and return an output.
Example: 

```{r}
y <- 10
class(y)
```

Pay attention to your data types, as they can cause unexpected behavior. For
example, sometimes numbers can be encoded as characters by R, which will result
in errors when you try to apply numerical operations to them. To create a
numerical vector, we can use the function `c()`:

```{r}
v <- c(1.2, 2.3, 1.1, 4.5, 6.3)
```

If you're not sure what a function does, you can check its documentation using
`help` or `?`:

```{r}
help(c)
?c
```

You can multiply a vector by a scalar, or do element-wise multiplication of
two vectors of the same length, using

```{r}
v*5
v2 <- c(0.3, 5.6, 1.4, 2.8, 3.5)
v*v2
```

Specific elements in a vector can be accessed or manipulated using indices
with square brackets: 

```{r}
v[2]
v[2:4]
v[2] <- 0.3
```

A variety of built-in functions exist to calculate basic summary statistics
from vectors: 

```{r}
max(v)
min(v)
sum(v)
mean(v)
median(v)
var(v)
sd(v)
```

Another key data structure in R is the *matrix*. Matrices are very common
in phylogenetics; input datasets for phylogenetic reconstruction are usually
in the form of a species by trait matrix, and they are also used extensively
in phylogenetic comparative methods (discussed more later in the course). 
There are generally two approaches to creating matrices in R. The first is 
using the function `cbind()` to concatenate existing vectors into columns.
Example using `v` and another vector of random numbers generated with
`rnorm`:

```{r}
v2 <- c(rnorm(5, mean = 2, sd = 5))
m <- cbind(v, v2)
dim(m) #Prints dimensions of matrix, rows x columns
```

You can also use `rbind` to bind by rows instead of columns.You can also create
a matrix directly using the `matrix()` function, passing a vector and specifying
the number of rows and columns: 

```{r}
v3 <- c(rnorm(12, mean = 2, sd = 5))
m2 <- matrix(v3, nrow = 3, ncol = 4)
```

You can access subsets of values in a matrix with indexing, using the 
general format `[rows,columns]`. Some examples:

```{r}
m2[1,] #First row of df
m2[,2] #Second column of df
m2[1:2,] #First and second rows
m2[,3:4] #Third and fourth columns
m2[,c(1,3)] #First and third columns
```

# Working with datasets

A *data frame* is a list of columns that can contain different
types of data. Most datasets that you will work with in R will be stored
as data frames, and many tools exist to extract and manipulate information
from them, both in base R (functions that come with R) and in additional
toolkits like `tidyverse`. Matrices can be converted into dataframes using
`as.data.frame()`:

```{r}
m2_df <- as.data.frame(m2)
#Add a new column to our dataframe identifying species
species <- c("species1", "species2", "species3")
m2_df <- cbind(species, m2_df)
colnames(m2_df) <- c("Species", "x", "y", "z", "a")
m2_df
```

Here, we've created a dataframe where three species have a measured value for
four different variables, x, y, z, and a. This data frame is in **wide** format;
each measured variable has its own column, and each row represents all the 
observations made for each species. Data frames can also be formatted **long**,
where each row represents a single observation of a single variable, creating
more rows and fewer columns. Many functions in R will require data to be 
passed in either wide or long formats, so learning how to change between them
is helpful. Here, we will work through an example using the R package `tidyr`,
which is part of `tidyverse`. First, we need to check if `tidyverse` is 
installed, and install it if not:

```{r}
#install.packages("tidyverse")
library(tidyr) #If this throws an error, uncomment the previous line to install
```

```{r}
#Example with pivot_longer here
m2_df_long <- m2_df %>%
                pivot_longer(!Species, names_to = "trait", values_to = "value")
m2_df_long
```

In the above code snippet, we made use of the tidyverse pipe, `%>%`. This 
passes the object preceding the pipe as the first argument to the function
after the pipe, and can often be used to write simpler and more readable code. 
Alternatively, the wide dataframe `m2_df` could have been passed directly
to the first argument of `pivot_longer()`. We converted the four continuous 
character values, x, y, z, and a, to a single variable called `trait` with
four possible states. 

Most data analysis projects in R consist of applying functions implemented in
various packages to data frames. Base R comes with many functions for basic
statistical analyses, such as the function `lm()` for linear regression 
analyses. Often we must search for the best package/function for the job,
and read its documentation to get an idea of how to run it. 

In next week's lab, you will learn about basic data structures for phylogenetic
analyses in R, tools for tree visualization, and applying basic manipulations
to tree data, including pruning, merging, and renaming nodes / tips. For now,
to receive credit for this week's assignment, knit this Rmarkdown file to
an html or pdf file, and upload the generated file with all code outputs 
to the assignment repository created for you in the class GitHub organization. 