---
title: "BIOL266_Lab3_Assignment.Rmd"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
remove(list=ls())
```

In this lab, you will learn about the basic data structures used to
store phylogenetic trees, some basic manipulations of those data
structures in R, and popular R packages for phylogenetic analyses. 

The common format used to store phylogenetic trees across languages and 
platforms is called a *Newick string*. In a Newick string, the hierarchical
structure of phylogenies is represented using a nested set of parentheses, with
each set of closed parentheses representing a clade. Commas are used to 
represent individual branches that descend from a single node. For example,
a phylogeny where species A and B are sister, C is sister to A/B, and D is 
sister to A/B/C would have the Newick string `(((A,B),C),D);`. A semicolon
is used to indicate the end of the string. 

Consider the following three Newick strings: 

```
(D,(C,(A,B)));
(((A,B),C),D);
(((B,A),C),D);
```
Do these strings represent the same or different topologies? Review material
from lecture 2 if necessary. Explain your reasoning in the box below:

```

```

Now consider the string `((A,B),C,D);`. What makes this string different
from the previous three? Explain below: 

```

```

In Newick format, the length of the branch descending from each node
is indicated by a semicolon `:` followed by a numerical float indicating the
length value. For tip branches, the semicolon is placed after the taxon name
for the branch; for internal branches, it is placed after the close parenthesis
indicating the clade the internal branch leads to. For example, the length of
the internal branch leading to the ancestor of `A` and `B` would be written
as `(A,B):0.01`, and the length of the tip branch leading to species C would
be written simply as `C:0.01`. Consider the following two strings:

```
(D:3,(C:2,(A:1,B:1):1):2);
(D:0.025,(C:0.01,(A:0.003,B:0.004):0.005):0.008);
```
Recall the difference between a chronogram and a phylogram as explained in
lecture 2. Which would you say best describes each of these trees? Explain
your reasoning below:

```

```

Consider a phylogeny where X is sister to Y, Z is sister to W, the XY clade
is sister to the ZW clade, and O is sister to the clade containing X, Y, Z, 
and W. Write this phylogeny in Newick format in the space below: 

```

```

To work with phylogenies in R, the Newick format needs to be converted into
a data structure that's R-compatible. This is typically an object of class
`phylo`, which is a list with a particular structure. Most R packages for 
phylogenetics have functions which can read in Newick strings and store the
resulting trees as `phylo` objects. For this assignment we will be using
`ape` (Analyses of Phylogenetics and Evolution). If you don't have this package
installed already, you can do so and load it with the chunk below: 

```{r}
#install.packages("ape") #Uncomment this if you need to install
library(ape)
```

Let's load in one of our example Newick strings from earlier using the function
`read.tree()`:

```{r}
tree <- ape::read.tree(text = "(D:0.025,(C:0.01,(A:0.003,B:0.004):0.005):0.008);")
tree
```

In addition to the `text` argument used here, you can also pass the function
a file with a Newick string on the first line. Note you can also write from
`phylo` object back to Newick string using `ape::write.tree()`. You can plot 
the resulting tree using `plot()`:

```{r}
plot(tree)
```

Let's have a look at the structure of the resulting phylo object: 

```{r}
str(tree)
```

In the space below, explain the `edge.length`, `Nnode`, and `tip.label` 
attributes, and what parts of the phylogeny they represent: 

```

```

The `edge` attribute is what encodes the tree topology, and it is a little 
less intuitive. Let's look at it more closely: 

```{r}
tree$edge
```

Each row represents an *edge*, or branch, in the tree, which
connects a parent node (left column) to a daughter node (right column). The 
order of rows is determined by a *preorder traversal*; each edge is visited by
starting from the root and taking each unique path from the root to the tips. 
By convention, the root is labelled as n+1 (where n is the number of tips) and
subsequent internal nodes increase from there by order in which they are 
visited, up to `n + Node`. Tips begin at 1 and also increase by the order 
in which they are visited, ending at `1 + Nnode`. In the box below, 
explain verbally which node is represented by each number in `tree$edge`:

```

```

Packages like `ape` come with a variety of functions for manipulating and 
extracting information from trees. For these exercises, we'll work with an
empirical phylogeny of Darwin's finches from the package `geiger`:

```{r}
#install.packages("geiger") #Uncomment if you need to install geiger
library(geiger)
data(geospiza)
geospiza_tree <- geospiza$phy
```

For each of the exercises below, rather than providing you with code,
I will encourage you to look through the documentation of `ape` (or other
packages if you prefer!) to find functions relevant to the task. First,
prune the tree of the taxa `pauper`, `difficilis`, and `fortis`, save the new
tree as `geospiza_tree_pruned`, and replot the pruned tree: 

```{r}

```

From the original tree, extract a subtree for the clade containing `difficilis`
and `fluginosa`, and plot it below: 

```{r}

```

Check if the following two sets of taxa form a monophyletic group in the
original tree: `Platyspiza, pauper, psittacula, parvulus`, `conirostris, magnirostris, fortis, fuliginosa`

```{r}

```

Get the number assigned to the node that is the most recent common ancestor
of `pauper` and `scandens`. What clade does this node define, and how is it 
related to the rest of the phylogeny? 

```{r}

```

Re-root the tree on `Pinaroloxias`. How does this change the perceived order
of branching events? 

```{r}

```

Finally, check if the tree is ultrametric:

```{r}

```